\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2017-11-22 document/XSGRB_sample.tex
Adding continuum estimate section

The {\it Swift} satellite \citep{Gehrels09}, which was launched in 2004, has
made it possible to harvest much of the great potential in using GRBs as probes
of galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.][]{Ricker04}.  A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based
follow-up observations of the afterglows and of the host galaxies of the GRBs.
This work has involved a large community of researchers both inside and outside
the official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs 
\citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability
to secure information for even very faint targets. This allows the measurements
of a number of important parameters such as redshifts, spectral slopes,
and extinction. For
a handful of very bright afterglows high-resolution (typically R$>$20000)
spectra were secured and for these events much more information about
conditions inside the host galaxies were extracted 
\citep[e.g.,][]{Fiore,Thoene07,Prochaska07,Vreejswijk07,CastroTirado2010}.

@@

We additionally provide an estimate of the continuum for all the spectra
presented here. For this, we have developed an algorithm that attempts to automatically
finds the continuum placement along with the error on the continuum estimate.

@@

All the spectra are made available at
\url{http://www.dark-cosmology.dk/~jselsing/XSGRB/} and additionally at , including the ones outside
the statistical sample. This release includes both prompt afterglow observations
as well as late time observations. An overview of all the spectra provided here
are given in Table. \ref{sample_overview}. For each burst, each individual
observation is provided in a separate reduction and in cases where observations
have been repeated, a combined spectrum is also provided.  No attempt has been
made to combine the arms, so for each observation, three spectra exists.

@@

@@

\begin{figure*}
	\centering
	\includegraphics[width=16cm]{figures/XSGRB_redshift.pdf}
	\caption{Redshift distribution of the XSGRB sample.}
	\label{fig:z}
\end{figure*}

@@

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectographs. The most striking result found is that GRBs
typically are located behind very large column densities of neutral hydrogen.
Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars have a handful such systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified \citep[e.g., five
in][]{Noterdaeme2012}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep[][and this work]{Jakobsson2006,
	Fynbo2009,Cucchiara2015}.  With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we hence characterize the properties of
star-forming galaxies over cosmic history in terms of redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, extinction curves,
etc.  A number of independent papers have been published or submitted for
publication focusing on many of these specific issues of our sample such as
extinction curves [][see also
\cite{Fynbo2014,Heintz2017}]\citep{Japelj2015,Zafar17}, emission lines from the
underlying host galaxies \citep{Kruhler2015}, the frequency of intervening
\ion{Mg}{ii} absorbers \citep{Christensen2017}, and escape of ionizing
radiation \citep{Tanvir17}.

It seems important to end this paper by stressing that the potential of 
using GRB sightlines as probes is far from fully harvested. The sample of
sightlines probed by our spectra is not representative for all GRB 
sightlines as we have shown and consistent with earlier findings
from samples based on low-resolution spectroscopy \citep[e.g.,][]{Fynbo2009}
and from studies of complete samples of GRB host galaxies 
\citep{Hjorth2012,Covinto2013,Perley2016}. 
\cite{Kruhler2013} convincingly argue, that 
very rich sightlines like that probed by the remarkable GRB\,080607 
\citep{Prochaska09} are probably significantly more frequent than in the sightlines 
sampled by our spectra. However, with current instrumentation these sightlines are
out of reach except under very fortunate circumstances as in the case of 
GRB\,080607 when the afterglow could be observed only a few minutes after the 
burst with a 10-m class telescope. Observations of such sightlines with 
X-shooter-like spectrographs on th enext generation of 20--40-m telescopes
is likely to be very rewarding.

@@

\subsection{GRB120119A (z = 1.728)}
The data presented here has not been published before. Three epochs of
observations have been obtained, the first two immediately after the burst, and
the last one long after the afterglow had faded. Starting 1.4 hours after the
\textit{Swift} trigger, the first epoch contains bright afterglow continuum.
Rich in absorption features belonging to a multitude of ions, $z =	1.728$ is
estimated for the host with intervening systems at $z =	1.476$, $z = 1.214$, $z
= 0.662$ and $z = 0.632$. The second epoch, obtained 4.5 hours after the burst
contains the fading afterglow. A third epoch is obtained $>1$ year after the
GRB in which emission lines from \hb~and \ha~are found at the redshift of the
host, confirming the association of the absorption line system and the host.

@@

\subsection{GRB140506A  (z = 0.889)}

@@

\subsection{GRB160804A (z = 0.736)}
The data presented here also formed the basis of GCN \#
19773\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/19773.gcn3}}, but is not
published elsewhere. Observations started 22.37 hours after the BAT trigger and
lasted for 2.4ks. The afterglow continuum is detected across the entire
spectral coverage of X-shooter and absorption lines from \mgi, \mgii, \feii~and
\alii~are found at $z = 0.736$. At the same redshift, emission lines from \oii,
\oiii, \ha, \hb, \hg, \nii, \sii, \siii~are found. A second epoch, lasting
3.6ks, is obtained after the afterglow has faded, confirming the emission line
detections.

>>>>

The {\it Swift} satellite \citep{Gehrels2009}, which was launched in 2004, has
made it possible to harvest much of the great potential in using GRBs as probes
of galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.][]{Ricker2004}.  A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based
follow-up observations of the afterglows and of the host galaxies of the GRBs.
This work has involved a large community of researchers both inside and outside
the official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs 
\citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability
to secure information for even very faint targets. This allows the measurements
of a number of important parameters such as redshifts, spectral slopes,
and extinction. For
a handful of very bright afterglows high-resolution (typically R$>$20000)
spectra were secured and for these events much more information about
conditions inside the host galaxies were extracted 
\citep[e.g.,][]{Fiore2005, Thone2007, Prochaska2007, Vreeswijk2007, Castro-Tirado2010}.

@@

We additionally provide an estimate of the continuum for all the spectra
presented here. For this, we have developed an algorithmic approach that
attempts to automatically finds the continuum placement along with the error on
the continuum estimate through an iterative procedure. The method is entirely
data-driven and does not rely on any physical assumptions. The method is used on
each arm separately for each spectrum.

To estimate the continuum, a number of points (typically on the order of 100)
are inserted at random positions along the wavelength direction, the
flux-density of each point determined by the median value of the spectrum in a
small region ($\sim 1 \mathrm{\AA}$) surrounding each point. The points are
fitted with a low order polynomial (we use \texttt{numpy.polynomial.chebyshev})
and iteratively, the point furthest away are removed until the polynomial fit
and points differ by less than $\sim$ 5 per cent. This filtering is used because
the afterglow continuum of GRBs are power laws \citep{Piran2005}, and by
removing points that differ significantly from a smooth continuum shape will
guide the continuum estimate to a shape more reasonable for GRB afterglows.
Additionally, points spaced less than around 1 per cent of the entire spectral
coverage are pruned. The remaining filtered and pruned points are then spline
interpolated using \texttt{scipy.interpolate.splrep;splev}, which serves as a
first estimate of the continuum placement. An attempt to identify absorption and
emission regions are then carried out, where they are marked as such if the
difference between the estimated continuum and the observed spectrum is larger
than 3 - 5 times the associated error spectrum. All regions marked as affected are then masked.

Using the masked spectrum, the entire process is then repeated 500 times where
the final continuum estimate is the mean of the continuum realizations and the
associated error estimate is the standard deviation. This error reflects the
stability of the algorithm across the spectrum. An example of the performance is
shown in Fig. \ref{fig:spectrum}. In some cases, very little continuum is
contained in a single arm and a manual continuum estimate is provided, similar
to what is done in \citet{Lopez2016}. In these cased the continuum error is set
to 10 per cent. The code for the continuum estimate is released along with the
paper at \url{https://github.com/jselsing/XSGRB-sample-paper}.

The continuum estimate and the associated error is distributed in a separate
extension, as described in Sect. \ref{products}.

@@

All the spectra are made available at
\url{http://www.dark-cosmology.dk/~jselsing/XSGRB} and additionally at , including the ones outside
the statistical sample. This release includes both prompt afterglow observations
as well as late time observations. An overview of all the spectra provided here
are given in Table. \ref{sample_overview}. For each burst, each individual
observation is provided in a separate reduction and in cases where observations
have been repeated, a combined spectrum is also provided.  No attempt has been
made to join the arms, so for each observation, three spectra exists.

@@

% Text for redshift distribution section
One of the primary objectives of our follow-up campaign is to measure the
redshift distribution for a well-defined, observationally unbiased and
statistical useful sample of GRBs. The imposed selection criteria (see Sect.
\ref{samplecrit}) ensures that the GRBs entering our homogenous sample is a good
representation of the underlying population. The redshift distribution of such
a sample holds valuable information about the occurance of GRBs through cosmic
time, which can then be directly compared to the star-formation rate inferred
from luminosity-selected samples of field galaxies.

In Fig.~\ref{fig:z} we show the redshift distribution of all GRBs observed with
X-shooter and for all the GRBs entering our sample. In the top panel we show a
histogram of the two distributions and below we show the redshifts of the
individual bursts as a function of GRB luminosity in the observed 15-150 keV
band. To calculate the luminosity, $L_{\mathrm{BAT}}$, we follow a similar
procedure as Lien(+16). We define this quantity as $L_{\mathrm{BAT}} =
F_{\gamma}\,4 \pi\,d_L^2\,(1+z)^{-1}$, where $F_{\gamma}$ is the observed BAT
fluence in the 15-150 keV band and $d_L$ is the luminosity distance to the
burst at the given redshift. We clearly see a correlation between the minimum
luminosity of the bursts and the redshift, mainly due to the Malmquist bias.
The lower limit in luminosity of our sample as a function of redshift
corresponds to a sensitivity of $10^{-7}$ erg cm$^{-2}$ s$^{-1}$ in the
observed 15-150 keV band (contrary to the expected value of $10^{-8}$ erg
cm$^{-2}$ s$^{-1}$; Swift/BAT+05). This is likely due to the more complex
detection criteria defined by Swift \textbf{(expand by Daniele?)}. Due to the Malmquist
bias the Swift luminosity detection limit is almost an order of magnitude
brighter at $z=2$ than at $z=1$. At $z\geq4$ we are only able to observe GRBs
that are hundred times brighter than the faintest bursts at $z=1$ and below.

Except for the two intrinsically faintest bursts at $z\approx 1 - 1.5$, the
statistical sample follows a constant lower limit in observed fluence as a
function of redshift. This means that our selection is not biased at high
redshifts except for the luminosity bias invoked by Swift. We find the
following values for the redshift distribution. First for the statistical
sample: $z_{\mathrm{avg}}=1.99$, $z_{\mathrm{med}}=1.64$, $z_{14\%}=0.60$ and
$z_{86\%}=3.39$. For all GRBs observed with X-shooter we find
$z_{\mathrm{avg}}=1.90$, $z_{\mathrm{med}}=1.52$, $z_{14\%}=0.55$ and
$z_{86\%}=3.36$.

@@

\begin{figure*}
	\centering
	\includegraphics[width=16cm]{figures/XSGRB_zBATlum.pdf}
	\caption{Redshift distribution as a function of intrinsic BAT $\gamma$-ray
		luminosity, $L_{\gamma,\mathrm{int}}$. The latter quantity is defined as
		$L_{\gamma,\mathrm{int}} = F_{\gamma,\mathrm{obs}}\,4 \pi\,d_L^2\,(1+z)^{-1}$,
		where $F_{\gamma,\mathrm{obs}}$ is the observed BAT fluence in the 15-150 keV
		band and $d_L$ is the luminosity distance at the given redshift. Blue star
		symbols denote the statistical sample and black dots show all GRBs observed
		with X-shooter and by \textit{Swift}/BAT. Overplotted as the red star symbols
		for comparison are the GRB, quasar, and galaxy with the highest
		spectroscopically confirmed redshifts, the latter two shown at arbitrary
		$L_{\gamma,\mathrm{int}}$. The blue solid line represents the BAT sensitivity
		of $10^{-8}$ erg cm$^{-2}$ s$^{-1}$ in the 15-150 keV band. The blue dashed
		line marks an observed sensitivity of $10^{-7}$ erg cm$^{-2}$ s$^{-1}$, which
		represents our sample limit out to $z\approx 8$. The estimated epoch of
		reionization is shown by the black dot-dashed line, with the uncertainty shown
		as the black dashed lines, from the most recent measurement by Planck+16.}
	\label{fig:z}
\end{figure*}

@@

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectographs. The most striking result found is that GRBs
typically are located behind very large column densities of neutral hydrogen.
Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars have a handful such systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified \citep[e.g., five
in][]{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep[][and this work]{Jakobsson2006b,
	Fynbo2009, Cucchiara2015}.  With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we hence characterize the properties of
star-forming galaxies over cosmic history in terms of redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, extinction curves,
etc.  A number of independent papers have been published or submitted for
publication focusing on many of these specific issues of our sample such as
extinction curves [][see also
\cite{Fynbo2014, Heintz2017}]\citep{Japelj2015, Zafar2017}, emission lines from the
underlying host galaxies \citep{Kruhler2015}, the frequency of intervening
\ion{Mg}{ii} absorbers \citep{Christensen2017}, and escape of ionizing
radiation \citep{Tanvir2017}.

It seems important to end this paper by stressing that the potential of 
using GRB sightlines as probes is far from fully harvested. The sample of
sightlines probed by our spectra is not representative for all GRB 
sightlines as we have shown and consistent with earlier findings
from samples based on low-resolution spectroscopy \citep[e.g.,][]{Fynbo2009}
and from studies of complete samples of GRB host galaxies 
\citep{Hjorth2012, Covino2013, Perley2016}. 
\cite{Kruhler2013} convincingly argue, that 
very rich sightlines like that probed by the remarkable GRB\,080607 
\citep{Prochaska2009} are probably significantly more frequent than in the sightlines 
sampled by our spectra. However, with current instrumentation these sightlines are
out of reach except under very fortunate circumstances as in the case of 
GRB\,080607 when the afterglow could be observed only a few minutes after the 
burst with a 10-m class telescope. Observations of such sightlines with 
X-shooter-like spectrographs on th enext generation of 20--40-m telescopes
is likely to be very rewarding.

@@

\subsection{GRB120119A (z = 1.728)}
The data presented here has been examined by Japelj+15 and Zafar+17 who both 
find a significant amount of extinction, $A_V\approx 1$ mag. Three epochs of
observations have been obtained, the first two immediately after the burst, and
the last one long after the afterglow had faded. Starting 1.4 hours after the
\textit{Swift} trigger, the first epoch contains bright afterglow continuum.
Rich in absorption features belonging to a multitude of ions at $z =    1.728$ is
estimated for the host with intervening systems at $z = 1.476$, $z = 1.214$, $z
= 0.662$ and $z = 0.632$. The second epoch, obtained 4.5 hours after the burst
contains the fading afterglow. A third epoch is obtained $>1$ year after the
GRB in which emission lines from \hb~and \ha~are found at the redshift of the
host, confirming the association of the absorption line system and the host. We
also detect C\,\textsc{i} in absorption which indicates the presence of cold gas.

@@

\subsection{GRB140506A  (z = 0.889)}	
The data presented here has formed the basis of GCN \#
16214\footnote{\url{http://gcn.gsfc.nasa.gov/other/140506A.gcn3}} and is
published in Fynbo et al. (2014) and Heintz et al. (2017a). The observations
consists of $4\times 600$ sec at 8.8 and 33 hours after the burst. We detect
absorption features from a wide range of ions, together with molecular
absorption from CH+, all at $z=0.889$. The optical/near-infrared afterglow
reveal an unusual steep extinction curve which is found to be caused by dust
immediate to the burst.

@@

\subsection{GRB160804A (z = 0.736)}
The data presented here also formed the basis of GCN \#
19773\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/19773.gcn3}}, and is
published in Heintz et al. (2017b). Observations started 22.37 hours after the
BAT trigger and lasted for 2.4ks. The afterglow continuum is detected across the entire
spectral coverage of X-shooter and absorption lines from \mgi, \mgii, \feii~and
\alii~are found at $z = 0.736$. At the same redshift, emission lines from \oii,
\oiii, \ha, \hb, \hg, \nii, \sii, \siii~are found. A second epoch, lasting
3.6ks, is obtained after the afterglow has faded, confirming the emission line
detections. The host galaxy is found to have a roughly solar metallicity and is
among the most luminous GRB hosts at $z < 1$.
